Lattice Boltzmann simulations of the dynamics of thin liquid films \\ \large{Gitter Boltzmann Simulationen der Dynamik von d\"unnen Fl\"ussigkeitsfilmen}

### 2024-01-30 chapters/Conclusion.tex
Printed and evaluated hopefully

To this end a novel approach for modelling the dynamics of thin liquid films has been developed, see Chap.~\ref{chapter:theory}-\ref{chapter:second_paper}.
It is important to highlight that this model consists of two parts. 
There is the theoretical work, where we use shallow water theory and find matching conditions that allow us to solve the thin film equation. 
In Chap.~\ref{chapter:theory} we introduce both the shallow water theory and the thin film equation and discuss some overlap between them, we however like to point out that we are not the first ones to do so.  
What makes our approach novel is the step from the continuum to a discrete system. 
Instead of solving the (stiff) degenerate fourth order partial differential equation for the evolution of the film thickness we solve a modified shallow water system using the lattice Boltzmann method.
A very thorough referee called this approach hyperbolization of the thin film equation.
The second part is therefore the numerical tool or solver that we created during this project.

To summarize the results of this thesis, we should first address the elephant in the room: ``Do we need yet another solver for the thin film equation?''
Although it might be tempting to answer this question with a clear ``Yes'' or ``No'', it is more complicated than that.
Computational fluid dynamics is a rich scientific field populated with several methods build on truly stunning results.
It is not just the finite volume method or the lattice Boltzmann method, but a vast collection of different approaches, many more than mentioned in this work.
There are various reasons why there are so many tools doing essentially the same. 
Some reasons grew historically, others developed with advances in mathematics.
One thing to mention is that it is of scientific value when a new method can reproduce known results.
That is because all methods suffer from shortcomings and there is no single method that is better suited in all hydrodynamic regimes than another one.
In Chap.~\ref{chapter:method} we introduced the idea behind the Chapman-Enskog expansion and found that the lattice Boltzmann method best works in the low $Kn \& Ma$ number regime.
Finite volume methods are highly dependent on the quality of the mesh and developing self-consistent boundary conditions can be a tedious task, e.g., contact angle dynamics.
Molecular dynamics simulations are inherently noisy and computationally demanding for large volumes, and yet all of those three methods can be used to simulate thin film problems.
Each method can address different aspects of a problem and yet all have to satisfy the underlying physical boundary conditions, e.g., mass conservation.
We briefly addressed why our method fits well into the current landscape of thin film solvers in Sec~\ref{section:statement_software} of Chap.~\ref{chapter:fourth_paper}.

@@

Picking up the question from above, ``Do we need yet another solver...~?'', we found some arguments for the ``No'' answer, let us discuss some arguments for the ``Yes'' answer.
One rather unscientific strength of this solver is the transparent nature of the software. 
The code base is available under an open source licence and everyone interested can fork the repository and use (and develop) our method.
It is becoming harder and harder to validate numerical simulations and argue about data, we think that transparency of our source code is at least a baby step into the right direction. 
The package approach, similar to e.g., Python, Matlab or R, makes it simple to share and collaborate with colleagues.
Apart from these quality of life/usability features there are few things that make the derived model special and scientifically interesting.
In contrast to various numerical methods used in the literature of thin film dynamics, the here presented method is build upon the lattice Boltzmann method as described in Chaps~\ref{chapter:method}-\ref{chapter:first_paper}.
It is therefore not intended as a general tool to approximate differential equations, but only for the hydrodynamics of thin liquid films.
While the term simplicity was used earlier as weakness, one of the nineteen guiding principles of python reads: ``Simple is better than complex''.
A small codebase is maintainable, it is possible to write tests for every function and add documentation, all of this would be harder for general purpose solvers with much more lines of code.
In Chap.~\ref{chapter:first_paper} we show that our simple approach is in fact more than capable to simulate relevant physical problems.
Multiple tests with analytical results have been used to validate the method, see Tanner's law and Cox-Voinov relation.
Another benefit of the here presented approach is the dimensional reduction.
While it is restricting on the one hand, e.g., contact angles smaller than $\pi/2$, it reduces the demand for computational resources quite significant.
That in fact allows us to introduce various additions to our model and still have reasonable fast running simulations.
All simulations presented in this thesis can be computed on a laptop with dedicated GPU, there is no real need for HPC resources.
Working on initial conditions and setting up parameters can be a tedious task that can make or break a simulation.
The here presented method turned out to be fairly robust and numerically stable in many regards.
To name just two, initial or ``boundary'' conditions with non-differentiable functions did not cause any issues, see e.g., Chap.~\ref{chapter:second_paper} Eq.~(\ref{eq:sharp_contact_angle_spatial}).
Although the method approximates the thin film equation and as such is only strictly valid for small contact angles, we found that contact angle up to $70^{\circ}$ seemed to be within the validity of the method, at least if we assume that Cox-Voinov is applicable in the regime.
Another feature that turned out to be of great value is the inclusion of hydrodynamic slip.
Having a method that can be used in the no-slip regime while also be applicable in the intermediate to large slip regime is outstanding.
And as shown in Chap.~\ref{chapter:second_paper} slip does in fact play an important role in the dynamics of a dewetting thin film.
While the model does not account for e.g., surfactants, we supply a matching condition between our method and thin film like theories in Chaps.~\ref{chapter:first_paper}-\ref{chapter:second_paper}.
The seemingly simple ``Yes'' or ``No'' question could be discussed in even more detail, but it is a subjective open question.
Therefore we would like to have a look at Chaps.\ref{chapter:second_paper}-\ref{chapter:third_paper} and their findings.

In Chap.~\ref{chapter:second_paper} we introduce thermal fluctuations to our model.
These fluctuations lead to a constant excitement of the fluid air interface that is smoothed out by the surface tension, leading to the appearance of thermo-capillary waves.
Their amplitude however is usually so small that they can be neglected in experimental analysis or numerical simulations. 
However having a dewetting thin film it can be assumed that the thickness of the film will be comparable to the amplitude of the fluctuations, at least in regions where the film ruptures.
As shown in Chap.~\ref{chapter:second_paper} these fluctuations can be added to the model using a matching condition for a fluctuating force term. 
The fluctuating force term is then tested against the theory of capillary waves and does show in fact good agreement with the predicted spectrum.
In the following it is shown that the stability difference, as measured in rupture times, between fluctuating and deterministic thin films~\footnote{$k_BT > 0$ fluctuating thin film, $k_BT = 0$ deterministic thin film.} does inversely depend on the logarithm of the wettability. 
By the virtue of the model a combination of a spatially varying wettability as well as the addition of thermal fluctuations is studied using a dewetting thin film.
Depending on the chosen function of the wettability it is possible to create numerical experiments where the deterministic simulation is almost indistinguishable from the fluctuating one.
At the beginning of this chapter we discuss the underlying theory and the modification to the thin film equation.
The constructive approach we use to match our model with the stochastic thin film equation should in fact serve as a blueprint for other additions e.g., surfactant dynamics.

In Chap.~\ref{chapter:third_paper} a simple toy model is developed to study dewetting dynamics under the influence of ``switchable'' substrates.
The model is build on the idea that the wettability, e.g., the equilibrium contact angle $\theta_{\text{eq.}}$, can be changed based on external stimuli. 
We assume that the these stimuli can be modelled by an arbitrary function.
In the here discussed case the arbitrary function is a trigonometric function ($\theta(x,t) \propto \sin(x t)$) that admits not only a spatial gradient but also a time dependent behaviour.
Due to this wettability gradient fluid is driven into regions of high wettability, or low contact angles.
Therefore the stationary state for vanishing time dependency, $\partial_t\theta = 0$, is simply an array of droplets formed in the minima of the contact angle field.
Adding a temporal component to the contact angle field introduces a stabilizing effect to the dewetting film, which lead to an increase in rupture time.
Most interestingly is the morphological transition that emerges when the dynamics of the contact angle field becomes fast as compared to the capillary retraction.
This means that the film finds itself in a metastable energy minimum where all fluid accumulates in rivulets.
Metastable because the rivulets are prone to an instability. 
In fact, independent of the parameters we vary rupture of these rivulets can be observed. 
The stationary state after the breakup is again a multi-droplet state, however with fewer droplets than in the static case.

>>>>

To this end a novel approach for modelling the dynamics of thin liquid films has been developed, see Chap.~\ref{chapter:theory} and the first three publications.
It is important to highlight that this model consists of two parts. 
There is the theoretical work, where we use shallow water theory and find matching conditions that allow us to solve the thin film equation. 
In Chap.~\ref{chapter:theory} we introduce both the shallow water theory and the thin film equation and discuss some overlap between them, we however like to point out that we are not the first ones to do so.  
What makes our approach novel is the step from the continuum to a discrete system. 
Instead of solving the (stiff) degenerate fourth order partial differential equation for the evolution of the film thickness we solve a modified shallow water system using the lattice Boltzmann method.
A very thorough referee called this approach hyperbolization of the thin film equation.
The second part is therefore the numerical tool or solver that we created during this project.

To summarize the results of this thesis, we should first address the elephant in the room: ``Do we need yet another solver for the thin film equation?''
Although it might be tempting to answer this question with a clear ``Yes'' or ``No'', it is more complicated than that.
Computational fluid dynamics is a rich scientific field populated with several methods build on truly stunning results.
It is not just the finite volume method or the lattice Boltzmann method, but a vast collection of different approaches, many more than mentioned in this work.
There are various reasons why there are so many tools doing essentially the same. 
Some reasons grew historically, others developed with advances in mathematics.
One thing to mention is that it is of scientific value when a new method can reproduce known results.
That is because all methods suffer from shortcomings and there is no single method that is better suited in all hydrodynamic regimes than another one.
In Chap.~\ref{chapter:method} we introduced the idea behind the Chapman-Enskog expansion and found that the lattice Boltzmann method best works in the low $Kn \& Ma$ number regime.
Finite volume methods are highly dependent on the quality of the mesh and developing self-consistent boundary conditions can be a tedious task, e.g., contact angle dynamics.
Molecular dynamics simulations are inherently noisy and computationally demanding for large volumes, and yet all of those three methods can be used to simulate thin film problems.
Each method can address different aspects of a problem and yet all have to satisfy the underlying physical boundary conditions, e.g., mass conservation.
We briefly addressed why our method fits well into the current landscape of thin film solvers in the first publication.

@@

Picking up the question from above, ``Do we need yet another solver...~?'', we found some arguments for the ``No'' answer, let us discuss some arguments for the ``Yes'' answer.
One rather unscientific strength of this solver is the transparent nature of the software. 
The code base is available under an open source licence and everyone interested can fork the repository and use (and develop) our method.
It is becoming harder and harder to validate numerical simulations and argue about data, we think that transparency of our source code is at least a baby step into the right direction. 
The package approach, similar to e.g., Python, Matlab or R, makes it simple to share and collaborate with colleagues.
Apart from these quality of life/usability features there are few things that make the derived model special and scientifically interesting.
In contrast to various numerical methods used in the literature of thin film dynamics, the here presented method is build upon the lattice Boltzmann method as described in Chaps~\ref{chapter:method} and the first two publications.
It is therefore not intended as a general tool to approximate differential equations, but only for the hydrodynamics of thin liquid films.
While the term simplicity was used earlier as weakness, one of the nineteen guiding principles of python reads: ``Simple is better than complex''.
A small codebase is maintainable, it is possible to write tests for every function and add documentation, all of this would be harder for general purpose solvers with much more lines of code.
In the second publication we show that our simple approach is in fact more than capable to simulate relevant physical problems.
Multiple tests with analytical results have been used to validate the method, see Tanner's law and Cox-Voinov relation.
Another benefit of the here presented approach is the dimensional reduction.
While it is restricting on the one hand, e.g., contact angles smaller than $\pi/2$, it reduces the demand for computational resources quite significant.
That in fact allows us to introduce various additions to our model and still have reasonable fast running simulations.
All simulations presented in this thesis can be computed on a laptop with dedicated GPU, there is no real need for HPC resources.
Working on initial conditions and setting up parameters can be a tedious task that can make or break a simulation.
The here presented method turned out to be fairly robust and numerically stable in many regards.
To name just two, initial or ``boundary'' conditions with non-differentiable functions did not cause any issues, see e.g.,  third publication.
Although the method approximates the thin film equation and as such is only strictly valid for small contact angles, we found that contact angle up to $70^{\circ}$ seemed to be within the validity of the method, at least if we assume that Cox-Voinov is applicable in the regime.
Another feature that turned out to be of great value is the inclusion of hydrodynamic slip.
Having a method that can be used in the no-slip regime while also be applicable in the intermediate to large slip regime is outstanding.
And as shown in the third publication slip does in fact play an important role in the dynamics of a dewetting thin film.
While the model does not account for e.g., surfactants, we supply a matching condition between our method and thin film like theories in the second and third publication.
The seemingly simple ``Yes'' or ``No'' question could be discussed in even more detail, but it is a subjective open question.
Therefore we would like to have a look at third and last publication and their findings.

In the third publication we introduce thermal fluctuations to our model.
These fluctuations lead to a constant excitement of the fluid air interface that is smoothed out by the surface tension, leading to the appearance of thermo-capillary waves.
Their amplitude however is usually so small that they can be neglected in experimental analysis or numerical simulations. 
However having a dewetting thin film it can be assumed that the thickness of the film will be comparable to the amplitude of the fluctuations, at least in regions where the film ruptures.
As shown in the third publication these fluctuations can be added to the model using a matching condition for a fluctuating force term. 
The fluctuating force term is then tested against the theory of capillary waves and does show in fact good agreement with the predicted spectrum.
In the following it is shown that the stability difference, as measured in rupture times, between fluctuating and deterministic thin films~\footnote{$k_BT > 0$ fluctuating thin film, $k_BT = 0$ deterministic thin film.} does inversely depend on the logarithm of the wettability. 
By the virtue of the model a combination of a spatially varying wettability as well as the addition of thermal fluctuations is studied using a dewetting thin film.
Depending on the chosen function of the wettability it is possible to create numerical experiments where the deterministic simulation is almost indistinguishable from the fluctuating one.
At the beginning of this chapter we discuss the underlying theory and the modification to the thin film equation.
The constructive approach we use to match our model with the stochastic thin film equation should in fact serve as a blueprint for other additions e.g., surfactant dynamics.

In the last publication a simple toy model is developed to study dewetting dynamics under the influence of ``switchable'' substrates.
The model is build on the idea that the wettability, e.g., the equilibrium contact angle $\theta_{\text{eq.}}$, can be changed based on external stimuli. 
We assume that the these stimuli can be modelled by an arbitrary function.
In the here discussed case the arbitrary function is a trigonometric function ($\theta(x,t) \propto \sin(x t)$) that admits not only a spatial gradient but also a time dependent behaviour.
Due to this wettability gradient fluid is driven into regions of high wettability, or low contact angles.
Therefore the stationary state for vanishing time dependency, $\partial_t\theta = 0$, is simply an array of droplets formed in the minima of the contact angle field.
Adding a temporal component to the contact angle field introduces a stabilizing effect to the dewetting film, which lead to an increase in rupture time.
Most interestingly is the morphological transition that emerges when the dynamics of the contact angle field becomes fast as compared to the capillary retraction.
This means that the film finds itself in a metastable energy minimum where all fluid accumulates in rivulets.
Metastable because the rivulets are prone to an instability. 
In fact, independent of the parameters we vary rupture of these rivulets can be observed. 
The stationary state after the breakup is again a multi-droplet state, however with fewer droplets than in the static case.
